---
title: "Predicting Readmission for Diabetes Inpatients"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 7, fig.height = 4)
if(!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr, ggplot2, glmnet, car, data.table, pROC)   #add your packages here
```

I. EDA

```{r}
diabetes <- read.csv('diabetic.data.csv', na.strings = "?")
readmission <- read.csv('readmission.csv', na.strings = "?")
```


```{r}
readmission$readmittedLess30 <- ifelse(readmission$readmitted == "<30", 1, 0)
readmission <- as.data.frame(unclass(readmission), stringsAsFactors = TRUE)
readmission$readmittedLess30 <- as.factor(readmission$readmittedLess30)
readmission <- subset(readmission, select = -c(readmitted))
```


```{r}
readmission <- na.omit(readmission)
sum(is.na(readmission))
glimpse(readmission)
#imputing the null values isnt worth the trouble becuase we have so much data
```
```{r}
#number of patients
length(unique(diabetes$patient_nbr))
#number of encounters
nrow(diabetes)
```

```{r}
summary(readmission)
```


```{r}
ggplot(readmission, aes(x = num_lab_procedures, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~disch_disp_modified)

ggplot(readmission, aes(x = num_medications, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~disch_disp_modified)


ggplot(readmission, aes(x = time_in_hospital, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~disch_disp_modified)

ggplot(readmission, aes(x = num_procedures , fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~disch_disp_modified)
 
     
ggplot(readmission, aes(x = number_outpatient, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~disch_disp_modified)

ggplot(readmission, aes(x = number_emergency, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~disch_disp_modified)

ggplot(readmission, aes(x = number_inpatient, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~disch_disp_modified)

ggplot(readmission, aes(x = number_diagnoses, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~disch_disp_modified)
```



```{r}
ggplot(readmission, aes(x = num_lab_procedures, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~age_mod)

ggplot(readmission, aes(x = num_medications, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~age_mod)


ggplot(readmission, aes(x = time_in_hospital, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~age_mod)

ggplot(readmission, aes(x = num_procedures , fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~age_mod)
 
     
ggplot(readmission, aes(x = number_outpatient, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~age_mod)

ggplot(readmission, aes(x = number_emergency, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~age_mod)

ggplot(readmission, aes(x = number_inpatient, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~age_mod)

ggplot(readmission, aes(x = number_diagnoses, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~age_mod)
```

```{r}
ggplot(readmission, aes(x = num_lab_procedures, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~insulin)

ggplot(readmission, aes(x = num_medications, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~insulin)


ggplot(readmission, aes(x = time_in_hospital, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~insulin)

ggplot(readmission, aes(x = num_procedures , fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~insulin)
 
     
ggplot(readmission, aes(x = number_outpatient, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~insulin)

ggplot(readmission, aes(x = number_emergency, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~insulin)

ggplot(readmission, aes(x = number_inpatient, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~insulin)

ggplot(readmission, aes(x = number_diagnoses, fill= readmittedLess30)) + geom_histogram(position = position_dodge()) +
theme_bw() + facet_wrap(~insulin)
```

II. Methodology and Model Selection

```{r}
#remove patient number and encounter id
readmission <- subset (readmission, select = -c(encounter_id,patient_nbr))
```

```{r}
#LASSO Prep
X <- model.matrix(readmittedLess30~., readmission)[,-1]
Y <- readmission[, 29]
dim(X)
colnames(X)
```

```{r}
set.seed(10)
fit1.cv <- cv.glmnet(X, Y, alpha=1, family="binomial", nfolds = 10, type.measure = "deviance")
plot(fit1.cv)
```

```{r}
coef.1se <- coef(fit1.cv, s="lambda.1se")
coef.1se <- coef.1se[which(coef.1se !=0),]
rownames(as.matrix(coef.1se))
```

```{r}
coef.min <- coef(fit1.cv, s="lambda.min")
coef.min <- coef.min[which(coef.min !=0), ]
rownames(as.matrix(coef.min))
```

We now compare between 1se and min. Fit.logit.1 uses 1se coefficients whereas fit.logit.2 uses min coefficients. The AUC for 1se model (fit.logit.1) is larger and it is more parsimonious so we select this model.

```{r}
fit.logit.1 <- glm(readmittedLess30 ~ time_in_hospital + num_medications + number_emergency + number_inpatient + number_diagnoses + insulin + diabetesMed + disch_disp_modified + age_mod + diag1_mod + diag3_mod, family = binomial, data = readmission)
```

```{r}
summary(fit.logit.1)
```

```{r}
Anova(fit.logit.1)
```

```{r}
fit.logit.2 <- glm(readmittedLess30 ~ num_medications + number_emergency + number_inpatient + number_diagnoses + insulin + diabetesMed + disch_disp_modified + age_mod + diag1_mod + diag3_mod, family = binomial, data = readmission)
```


```{r}
fit1.roc<- roc(readmission$readmittedLess30, fit.logit.1$fitted, plot=T, col="blue")
fit2.roc<- roc(readmission$readmittedLess30, fit.logit.2$fitted, plot=T, col="blue")
```


```{r}
# The AUC for 1se model (fit.logit.1) is larger and it is more parsimonious so we select this model.
auc(fit1.roc)
auc(fit2.roc)
```

III. Classifier

Mislabeling a readmission means false positive $a_{0,1}=L(Y=0, \hat Y=1)$
Mislabeling a non-readmission means false negative $a_{1,0}=L(Y=1, \hat Y=0)$
The cost of mislabeling a readmission is **twice** of that mislabeling a non-readmission
Which means $a_{0,1}=2a_{1,0}$

$$P(Y=1 \vert X) > \frac{\frac{a_{0,1}}{a_{1,0}}}{1 + \frac{a_{0,1}}{a_{1,0}}}$$
$$P(Y=1 \vert X) > \frac{\frac{2}{1}}{1 + \frac{2}{1}}$$
$$P(Y=1 \vert X) > \frac{2}{3}$$

```{r}
pred3 <- data.frame(p = seq(0,1, length.out=101))
```
```{r}
MCE_lst <- list()

for (p in pred3$p) {
  fit1.pred.b <- as.factor(ifelse(fit.logit.1$fitted > p, "1", "0"))
  MCE <- (sum(fit1.pred.b[readmission$readmittedLess30 == "1"] != "1")
              + 2* sum(fit1.pred.b[readmission$readmittedLess30 == "0"] != "0"))/length(readmission$readmittedLess30)
  MCE_lst <- c(MCE_lst, MCE)
} 
  
pred3$MCE <- as.numeric(MCE_lst)
```
```{r}
ggplot(pred3, aes(p, MCE))  + geom_line()
```

